Supplementary Material

### 2020-03-11 Revisions/Rebuttal.md

<!--
rmarkdown::render('./Revisions/Rebuttal.md')
rmarkdown::render('./Revisions/Rebuttal.md', 'word_document')
-->

<!-- =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= -->
# Letter to the editor
<!-- =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= -->

Dear Dr. Otto,

We are submitting a revised version of the manuscript “Next Generation Planning - Structuring and Sharing Environmental Drivers Data for the St. Lawrence System” (ID: 469411) for publication in Frontiers in Marine Science. We would like to thank the reviewer for very constructive comments. We feel that we were able to address all the comments and are now submitting an improved manuscript. Below, you will find our responses to the reviewer’s comments and how we have revised our manuscript in view of these.

We thank you for considering our revised manuscript and look forward to hearing back from you.

@@

Best regards,

David Beauchesne

# Responses to the reviewer’s comments

<!-- =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= -->
## Question 1
<!-- =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= -->

@@

***Please summarize the main findings of the study***

*This article presents 22 environmental drivers and uses these environmental drivers to identify areas of high cumulative exposure and characterize areas with similar cumulative exposure in the St. Lawrence System. In particular, the article introduces an open knowledge platform named eDrivers, though it doesn’t seem to be working at this moment (https://github.com/eDrivers - this site contains zero repositories, zero people, zero projects; the R package eDrivers is not available on CRAN). When this R package and the website become available, I believe the article and its related sources will be beneficial to the scientific and management communities. Despite its weaknesses (shown below), I believe this work is a good start for assessing and managing cumulative effects in the St. Lawrence System and can serve as a good example for other marine ecosystems.*

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
**Answer:**

> - The material has yet to be released publicly on GitHub as we were waiting for the review process to be completed. We can, however, provide access to the repositories for the review process if the reviewer wishes to review the contents.

> - As for the package and the data, it will be available on Github, but we will not release a version on CRAN just yet. The material is not yet implemented on the St. Lawrence Global Observatory (SLGO; https://ogsl.ca/en) servers (our partners on this project) and the data are currently embedded directly in the R package. Our goal is to publish the package and submit it for peer-review once the data are available and accessible through SLGO's API.

<!-- =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= -->
## Question 2
<!-- =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= -->

@@

***Comment 1:***

*Authors need to make it clear up in the front that this work is only focusing on one of the four elements required to evaluate environmental impacts. Besides moving Lines 119-122 to around Line 77, it will be helpful to introduce the best practices in cumulative effects assessment and management: (1) selecting valued (ecosystem) components, (2) selecting indicators, (3) setting temporal and spatial scales, (4) addressing uncertainties, and (5) public participation. It will also be helpful to differentiate the two different indicators: effects-based and stressor-based. Effects-based indicators are direct measurements of the valued components (such as species abundance, biodiversity, etc.) and inherently encompass cumulative effects of multiple drivers. However, because of the poorly understood cause-effect linkages between stressors and valued components, it is currently hard to be used for decision-making. In comparison, stressor-based indicators (i.e., driver-based as this article defines and the type that this article uses) are usually better understood and can be proactively linked to management actions. However, these indicators do not capture the effects of all stressors and are not able to discern non-additive cumulative effects (e.g. synergism, antagonism). This point leads to the next weakness of this article.*

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
***Answer:***

> - We agree with the reviewer about the importance of clearly stating that this manuscript solely focuses on drivers. The goal of this manuscript and the eDrivers initiative is to facilitate the widespread accessibility of such data for environmental assessments and integrative management. We believe that we are clearly stating that the goal of our manuscript is focused on a single element of a fully operational impact assessment given the title of the manuscript and the 3rd sentence of the abstract: “Our objective in this paper is to systematically characterize the distribution and intensity of drivers in the St. Lawrence System.” (lines 38-39). Our goal is not to address cumulative effects. It is rather to provide knowledge on an element common to most holistic approaches, i.e. drivers characterization. We strived to make this as clear as possible throughout the manuscript, namely in the introduction and in the perspectives sections, as noted by the reviewer. As such, and considering the focus of the manuscript, we are hesitant to accept the reviewer’s suggestion to discuss cumulative effect assessment, best practices and indicator types at the beginning of the introduction. We feel that, considering the objective of this manuscript, the beginning of the introduction should focus specifically on drivers.

> - We, nonetheless, agree with the reviewer’s suggestions and agree that a broader context is necessary to properly position our manuscript in the environmental assessment literature. We are, therefore, proposing the following modifications to the manuscript, as we hope that these will consolidate the essence of the suggestions made by the reviewer:

  > 1. We modified the introduction to broadly state the major steps that are generally undertaken in holistic environmental approaches (lines 101-105).

  > 2. We modified the statement of the objectives that appears at the end of the introduction to clearly state that we are specifically focusing on drivers (lines 130-136).

  > 3. We modified the paragraph in the perspective sections (lines 496-518) in which the importance of valued components is discussed. It now presents eDrivers as a suite of useful, stressor-based indicators that, nevertheless,  fail to capture effects on valued components and non-additive effects. We present effects-based indicators and discuss their importance for understanding cumulative effects. We then argue that both are required to properly diagnose and attribute causes of ecosystem change to stressors (Dubé 2003, Environmental Impact Assessment Review 23(6):723-745; Jones 2016, Environmental Reviews 24(2):187-204).

@@

***Comment 2:***

*Because of this inherent weakness of stressor-based indicators, the dealing and conclusion of this article related to driver interactions (Sections 3.4 and 4.2) are inadequate. How different stressors (or drivers) interact to create nonlinear, non-additive cumulative effects are dependent on a specific valued component of interest. This article simply assumes the summed intensity of two drivers as interactions, which is inaccurate and misleading. Consequently, the results based on the two drivers (hypoxia and fisheries) are not convincing (Section 4.2). How to create indicators that truly reflect negative impacts is essential and can be specific to a specific value component. For instance, temperature positive anomalies may have negative impact on certain species, it may not necessarily degrade ecosystem production. While the calculation of drivers intensity in Appendix is only one way of doing it, it is important for the computer package to provide users with flexibility to define their own indicators (i.e., one potential solution to the weakness).*

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
***Answer:***

> - We thank the reviewer for pointing out this obvious weakness of the manuscript. We now realize that, even though we did not want to mislead readers in thinking we were evaluating driver interactions, we were doing just that. As noted by the reviewer, driver co-occurrence provides no information on valued component-specific driver interactions and non-additive effects. That said, driver co-occurrence is necessary for interactions to exist, and therefore it is expected that interactions and non-additive effects will be most likely observed in areas where drivers overlap at high relative intensities. These areas delineate where the effects of both drivers are most likely to impact valued components simultaneously. In sections 3.4 and 4.2, we illustrate how a simple two-drivers case could be used to identify areas of such high relative intensity driver overlap. This example should serve as a stepping stone to the integrative cumulative footprint and hotspot indicators. As rightfully noted by the reviewer, however, the manuscript is misleading on the topic of driver interactions. We have therefore made the following modifications to the manuscript:

@@

  > 1. We removed objective 2 from the manuscript and mentions of driver interactions, except in the context of discussing the likelihood of observing interactions in areas of driver co-occurrence.

  > 2. The methodology and results/discussions sections on driver interactions were moved to sections 3.3 and 4.1, respectively, and renamed Driver co-occurrence. We then modified the methods section and provided a simplified example of driver co-occurrence (lines 194-204). We did not, however, modify the results/discussion section. The only mention of driver interactions was in the title itself. In the discussion, we merely discuss the areas of high intensity of driver co-occurrence and describe individual and combined adverse effects that these drivers may have on species.

  > 3. We clarified our stance on driver co-occurrence and risk of interactions in the introduction, noting that individual and interactive effects of both drivers are species- and community-dependent and that the co-occurrence of drivers in itself is not evidence of driver interaction. We then state that, assuming driver co-occurrence is a prerequisite for driver interactions, it expected that interactions will occur in areas where both drivers co-occur simultaneoustly (lines 93-96).

> - Ultimately, this portion of the manuscript should serve as a simple illustration to better understand the complete cumulative footprint indicators. In this sense, if the reviewer still feels that the proposed changes are insufficient, we are willing to remove these sections from the manuscript.

> - As for providing flexibility in indicator definition, it is a feature that we wish to provide in future iterations of the package and platform. We would, however, rather not include such enhancements in this manuscript, as they are more technical in nature and we feel that they should not prevent us from publishing this work. Instead, we present this idea as a perspective and of future work (lines 524-526), as suggested by the reviewer.

@@

***Comment 3:***

*Because of the paramount importance of defining valued components in the process of cumulative effects assessment and management, the absence of this element is the third weakness of this article. This weakness can potentially be overcome by adding “Future work”. In the future, similar work can be carried out to select and mapping valued components, which will allow explicit linkages between drivers and value components to be explored for better understanding cumulative effects.*

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
***Answer:***

> - We agree with the reviewer on the importance of valued components for cumulative effects assessments. Presenting a full cumulative impact assessment and defining valued components are, however, not the objectives of this manuscript or of the platform, nor is it to discuss cumulative impacts in particular. By themselves, we believe stressor-based indicators offer valuable insights into the dynamics of a system which can help managers to prioritize decision-making and scientists in target specific areas for in situ measurements for hypothesis testing, especially when data on valued components are lacking. Furthermore, the choices of valued components are fully dependent on the objective of a particular assessment or management scheme. In our experience, time and money are spent on recurring initiatives that seek to characterize sources of stress in a system for impacts assessments and integrative management. Our goal was therefore to build an initiative that would curb this approach and provide access to formatted and comparable knowledge on sources of stress that are typically used for impact assessments and integrative management. This integrative dataset can then be used and adapted to the need of a user to assess impacts on a single or multiple valued components of interest, or the user can identify gaps in knowledge in the platform and ultimately contribute to the knowledge contained in the platform. We were therefore very deliberate in our decision not to include and discuss specific valued components. This manuscript is part of a broader project that aims at evaluating cumulative impact on the ecological communities of the St. Lawrence. The description of drivers will thus be used on specific valued components in the St. Lawrence System.

> - We already had included a paragraph on the importance of valued components and their vulnerability to sources of stress. We now realize, however, that it was ineffectual to provide perspective and direction as to how eDrivers may be used in the future. We therefore extended this paragraph, also in response to the reviewer’s first comment, to better address the significance of valued components and the different types of indicators used for cumulative effects assessment (lines 496-518).
-
<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->

<!-- =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= -->
## Question 3
<!-- =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= -->

@@

*Line 229: add “on” before the analysis.*

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
> - This was modified in the text.
<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->

*Lines 327-328: fix language issue.*

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
> - We have modified the sentences on lines 327-328 (now lines 348-352). "Of particular concern are threat complexes 3 and 6, which are the two most exposed complexes in the St. Lawrence and are characterized by distinct cumulative exposure regimes between them, they capture most of the coastal and offshore hotspots identified in the St. Lawrence." now becomes "Threat complexes 3 and 6 capture most of the coastal and offshore hotspots and are the two most exposed complexes in the St. Lawrence."
<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->

@@

*Figure 5 caption: tp becomes to*

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
> - The figure caption was modified accordingly.
<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->

@@

*Please check Appendix carefully for language errors. In addition, Appendix can be tightened up.*

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
> - We have performed a thorough review of the material presented in the Appendix.

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->

<!-- =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= -->
## Question 4
<!-- =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= -->

@@

*The method on dealing with driver interactions is inadequate, and the resultant conclusions with regard to interactions are not convincing.*

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
***Answer:***

> - See answer to question 2
<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->

<!-- =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= -->
## Question 5
<!-- =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-= -->

@@

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->

*Are the data underlying the study available in either the article, supplement, or deposited in a repository? (Sequence/expression data, protein/molecule characterizations, annotations, and taxonomy data are required to be deposited in public repositories prior to publication) (yes/no/not applicable)
- No*

<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~  -->
***Answer:***

> - The formatted driver data will be available through the package and the raw data will be archived on Zenodo and accessible upon request to the authors. The archive is ready to be published, but we are waiting for the publication process to be completed so that the proper citation can be referenced with the material. Likewise, the workflows used for the first iteration of the platform will be released and archived on Zenodo with the paper publication.

### 2020-03-11 Revisions/Beauchesne-eDriversMS-SI.rmd
manuscript revisions to submit to Frontiers

Table S1. List of drivers currently available on *eDrivers* along with their respective acronym used in the figures in the supplementary material.

@@

Oceans are the largest reservoir\textcolor{blue}{s} and sink\textcolor{blue}{s} of atmospheric carbon dioxide ($CO_2$).
Its uptake increases seawater acidity and \textcolor{red}{\sout{the}} lowers saturation state with
respect to calcite ($\Omega_C$) and aragonite ($\Omega_A$), the two most common
$CaCO_3$ polymorphs constituting the shells and skeleton\textcolor{blue}{s} of many marine
organisms [@mucci2017]. This can have deleterious effects on carbonate-secreting
organisms (*e.g.* mollusks and crustaceans) and certain physiological processes
in non-calcifying organisms [@fabry2008; @kroeker2013].

When $\Omega_C$ or $\Omega_A$ decrease below 1, water becomes undersaturated and
corrosive to the skeletal minerals of carbonate-secreting organisms. Note that
if $\Omega_A$ $< 1$, the waters may still be supersaturated with respect to
calcite since it is 50% more soluble than aragonite. However, organisms have
wide ranging responses to $\Omega_C$ and $\Omega_A$ saturation state. For
example, most corals stop calcifying at $\Omega_A$ $< 2$ [@kleypas2006; @bove2019],
while other organisms may have adapted to precipitate $CaCO_3$ even when
seawater is undersaturated with respect to calcite [$\Omega_C$ $< 1$ or
$\Omega_A$ $< 0.65$; e.g. @uthicke2016].

We used $\Omega_{A}$ saturation state to characterize ocean acidification in the
bottom waters of the St. Lawrence. The data come from the Department of
Fisheries and Oceans' (DFO) Atlantic Zone Monitoring Program [AZMP; @galbraith2018]
surveys and DFO's multispecies surveys [@bourdages2018] collected in
August-September of 2017 [@starr2019].
\textcolor{blue}{$\Omega_{A}$, pH, dissolved oxygen ($O_2$) and pressure were sampled at 117
stations in the Estuary and Gulf of St. Lawrence in the summer of 2018. The
majority of sampling stations were located in the Southern Gulf ($n = 84$),
compared to the Estuary and Northern Gulf ($n = 33$).}

The carbonate chemistry was determined
through pH and \textcolor{blue}{total alkalinity} (TA) measurements. Samples for pH and \textcolor{red}{\sout{total alkalinity}} TA were
collected under \textcolor{red}{\sout{a}} bubble free and no head space conditions into 500 $mL$
borosilicate glass flasks. 250 $\mu L$ of \textcolor{red}{\sout{the}} saturated $HgCl_2$ solution was
added to sample, and processed following the "Guide to best practices for Ocean
CO2 Measurements" [@dickson2007].

@@

TA was determined by potentiometric titration in an open cell using an
automated Radiometer potentiometric titrator (Titrilab 865) and a pH combination
electrode (pHC2001) in a continuous titrant addition mode, an algorithm
specifically designed for shallow end-point detection [@dickson2007]. The dilute\textcolor{blue}{d}
HCl titrant 0.1M in a solution of NaCl of 0.6M was calibrated using CRM provided
by Professor Andrew G. Dickson.

The carbonate system parameters (including $\Omega_{A}$) were
corrected for *in situ* pressure and temperature using the algorithm CO2SYS
[@lewis1998] with measured pH, \textcolor{red}{\sout{total alkalinity}} \textcolor{blue}{TA}, soluble reactive
and silicate concentrations as input parameters.

\textcolor{red}{\sout{$\Omega_{A}$, pH, dissolved oxygen ($O_2$) and pressure were sampled at 117
stations in the Estuary and Gulf of St. Lawrence in the summer of 2018. The
majority of sampling stations were located in the Southern Gulf ($n = 84$),
compared to the Estuary and Northern Gulf ($n = 33$).}}

To account for low sample size in the Northern Gulf, we used the correlation
between $O_2$ and $CO_2$, which are linked through the stoichiometry of the
respiration reaction in waters that are isolated from the atmosphere [@mucci2011].
Metabolic $CO_2$ increases in water through biotic processes such microbial
respiration of organic matter [@mucci2011]. In deep waters, variations of
$\Omega_{A}$ are largely governed by the addition of metabolic $CO_2$, whereas
near the air-sea interface $O_2$ is replenished faster than $CO_2$ can escape
[@zeebe2001]. The impact of respiration on carbonate chemistry is highest in
hypoxic regions, where metabolic $CO_2$ accumulates to high concentrations
[@mucci2011].

We explored the correlation between $\Omega_{A}$ and $O_2$ in the data and found
no significant correlation at the scale of the whole St. Lawrence (*p-value*:
0.08; $R^2$: 0.03). Nevertheless, a correlation between $\Omega_A$ and $O_2$
exists in the Northern St. Lawrence (*p-value*: < 0.01; $R^2$: 0.49), but not
in the Southern St. Lawrence (*p-value*: 0.85; $R^2$: < 0.01). Furthermore, the
correlation observed in the north between $\Omega_A$ and $O_2$ is especially
strong when deep Channels (*Deep* model: *p-value*: < 0.01; $O_2$: 0.20; $R^2$:
0.97) are separated from shallower regions (*Shallow* model: *p-value*: < 0.01;
$O_2$: 0.07; $R^2$: 0.96), with $\Omega_A$ increasing more steeply with
increasing $O_2$ in deep Channels. This stronger correlation is likely caused
by the lack of gas exchange with the atmosphere or the surface mixed layer and
the importance of metabolic $CO_2$ in deep waters, some of which may have mean
ages of up to 20 years [@mucci2011]. In the Northern Gulf, the acidification is
following depth and distance from the Cabot Strait along the Laurentian,
Anticosti and Esquiman Channels due to the progressive oxygen depletion and
metabolic carbon dioxide accumulation [@mucci2011].

There are likely different dynamics at play in the Southern Gulf, such as
waters with lowered alkalinity and salinity caused by the St. Lawrence plume.
Such water masses \textcolor{red}{\sout{are}} have lower buffering capacities to decreasing pH and
$\Omega_{A}$. Also, the accumulation of metabolic carbon dioxide produced by
rapid oxic degradation of organic matter in the sandy
and permeable sediments of the Magdalen Shelf or through the St. Lawrence plume
could decrease $\Omega_{A}$ in the Southern Gulf (Starr and Chassé, *personal
communication*).

We thus divided the analyses between the Southern and Northern St. Lawrence.
The St. Lawrence was divided using the 350m isobath on the southern
slope of the Laurentian Channel, from the Cabot Strait to the tip of the Gaspé
Peninsula. $\Omega_A$ distribution in the Southern St. Lawrence was modeled
using the exponential kriging model.

We predicted $\Omega_A$ using $O_2$ for the Northern St. Lawrence. \textcolor{red}{\sout{Whereas}} \textcolor{blue}{While}
there are few published datasets on $\Omega_{A}$ in the St.
Lawrence, there is a robust time-series on oxygen concentration in the water
column from the Department of Fisheries and Oceans’ (DFO) Atlantic Zone
Monitoring Program [AZMP; @galbraith2018]. We used oxygen concentration data
collected between 2013 and 2017
[@blais2019] and interpolated oxygen % saturation using cokriging with depth as a
covariable, as done in [@dutil2011]. Bathymetric data come from @dutil2011
and have a 100 $km^2$ resolution. For more details, refer to the hypoxia \textcolor{blue}{data layer}
description. The *Deep* model was used to transform $O_2$ values to $\Omega_A$
in grid cells with depth > 350m, while the *Shallow* model was
applied to cells with depth < 350m.

Based on the wide range of $\Omega_A$ levels at which organism are negatively
affected (see above), we built an index of acidification stress ($A_s$)
ranging between 0 and 1 that slowly \textcolor{red}{\sout{begins increasing}} \textcolor{blue}{increases} as $\Omega_A$ $\approx$
$2$, increases more rapidly close to $\Omega_A$ $= 1$ and reaches its peak at
$\Omega_A$ $\approx$ $0.5$:

@@

The data used to characterize hypoxia \textcolor{red}{\sout{uses oxygen saturation in % and}} come from
the Department of Fisheries and Oceans’ (DFO) Atlantic Zone Monitoring Program
[AZMP; @galbraith2018] survey\textcolor{red}{\sout{s}} in late spring and fall, as well as from \textcolor{red}{\sout{annual}}
DFO’s \textcolor{blue}{annual} multispecies surveys for the northern Gulf in August and for the Magdalen
Shallows in September. We provide a brief summary of data and methods to
describe hypoxia in this document. For more details, refer to @blais2019.

>>>>

Table S1. List of drivers currently available on *eDrivers* along with their respective acronym used in the figures \textcolor{red}{\sout{in}} \textcolor{blue}{of} the supplementary material.

@@

Oceans are the largest reservoir\textcolor{blue}{s} and sink\textcolor{blue}{s} of atmospheric carbon dioxide ($CO_2$). Its uptake increases seawater acidity and \textcolor{red}{\sout{the}} lowers \textcolor{blue}{the} saturation state \textcolor{blue}{of waters} with respect to calcite ($\Omega_C$) and aragonite ($\Omega_A$), the two most common $CaCO_3$ polymorphs \textcolor{blue}{that constitute} \textcolor{red}{\sout{constituting}} the shells and skeleton\textcolor{blue}{s} of many marine organisms [@mucci2017]. \textcolor{blue}{Ocean acidification} \textcolor{red}{\sout{This}} can have deleterious effects on carbonate-secreting organisms (*e.g.* mollusks and crustaceans) and certain physiological processes in non-calcifying organisms [@fabry2008; @kroeker2013].

When $\Omega_C$ or $\Omega_A$ decrease below 1, water becomes undersaturated and corrosive to the skeletal minerals of carbonate-secreting organisms. Note that if $\Omega_A$ $< 1$, the waters may still be supersaturated with respect to calcite since it is 50% more soluble than aragonite. However, organisms have wide ranging responses to \textcolor{blue}{changes in $\Omega_C$ and $\Omega_A$} \textcolor{red}{\sout{saturation state}}. For example, most corals stop calcifying at $\Omega_A$ $< 2$ [@kleypas2006; @bove2019], while other organisms may have adapted to precipitate $CaCO_3$ even when seawater is undersaturated with respect to calcite [$\Omega_C$ $< 1$ or $\Omega_A$ $< 0.65$; e.g. @uthicke2016].

We used $\Omega_{A}$ \textcolor{red}{\sout{saturation state}} to characterize ocean acidification in the bottom waters of the St. Lawrence. The data come from the Department of Fisheries and Oceans' (DFO) Atlantic Zone Monitoring Program [AZMP; @galbraith2018] surveys and DFO's multispecies surveys [@bourdages2018] collected in August-September of 2017 [@starr2019]. \textcolor{blue}{$\Omega_{A}$, pH, and dissolved oxygen ($O_2$) were measured at 117 stations in the Estuary and Gulf of St. Lawrence in the summer of 2018. The majority of sampling stations were located in the Southern Gulf ($n = 84$), the Estuary and Northern Gulf ($n = 33$).}

The carbonate \textcolor{red}{\sout{chemistry}} \textcolor{blue}{speciation} was determined through pH and \textcolor{blue}{total alkalinity} (TA) measurements. Samples for pH and \textcolor{red}{\sout{total alkalinity}} TA were collected \textcolor{red}{\sout{under a bubble free and}} \textcolor{blue}{with} no head space \textcolor{red}{\sout{conditions}} in\textcolor{red}{\sout{to}} 500 $mL$ borosilicate glass flasks\textcolor{red}{\sout{.}}\textcolor{blue}{, and} 250 $\mu L$ of \textcolor{red}{\sout{the}} saturated $HgCl_2$ solution was
added to \textcolor{blue}{each} sample\textcolor{red}{\sout{,}} and processed following the "Guide to best practices for Ocean
CO2 Measurements" [@dickson2007].

@@

TA was determined by potentiometric titration in an open cell using an
automated Radiometer potentiometric titrator (Titrilab 865) and a pH combination
electrode (pHC2001) in a continuous titrant addition mode, an algorithm
specifically designed for shallow end-point detection [@dickson2007]. The
\textcolor{red}{\sout{dilute}} \textcolor{blue}{0.1M} HCl titrant
\textcolor{red}{\sout{0.1M}} in \textcolor{red}{\sout{a solution of}}
\textcolor{blue}{0.6M} NaCl \textcolor{red}{\sout{of 0.6M}} was calibrated using CRM provided by Professor Andrew G. Dickson.

The carbonate system parameters (including $\Omega_{A}$) were
corrected for *in situ* pressure and temperature using the algorithm CO2SYS
[@lewis1998] with measured pH, \textcolor{red}{\sout{total alkalinity}} \textcolor{blue}{TA}, soluble reactive \textcolor{blue}{phosphate}
and silicate concentrations as input parameters.
\textcolor{red}{\sout{$\Omega_{A}$, pH, dissolved oxygen ($O_2$) and pressure were sampled at 117
stations in the Estuary and Gulf of St. Lawrence in the summer of 2018. The
majority of sampling stations were located in the Southern Gulf ($n = 84$),
compared to the Estuary and Northern Gulf ($n = 33$).}}

To account for \textcolor{blue}{the} low sample size in the Northern Gulf, we used the correlation
between $O_2$ and $CO_2$, which are linked through the stoichiometry of the
respiration reaction in waters that are isolated from the atmosphere [@mucci2011].
Metabolic $CO_2$ increases in water through biotic processes such microbial
respiration of organic matter [@mucci2011]. In deep waters\textcolor{blue}{, found in the Laurentian, Anticosti and Esquiman Channels}, variations of
$\Omega_{A}$ are largely governed by the addition of metabolic $CO_2$, whereas\textcolor{blue}{,} near the air-sea interface\textcolor{blue}{,} $O_2$ is replenished faster than $CO_2$ can escape
[@zeebe2001]. The impact of respiration on carbonate chemistry is highest in
hypoxic regions, where metabolic $CO_2$ accumulates to high concentrations
[@mucci2011].

We explored the correlation between $\Omega_{A}$ and $O_2$ in the data and found
no significant correlation at the scale of the whole St. Lawrence (*p-value*:
0.08; $R^2$: 0.03). Nevertheless, a correlation between $\Omega_A$ and $O_2$
exists in the Northern St. Lawrence (*p-value*: < 0.01; $R^2$: 0.49), but not
in the Southern St. Lawrence (*p-value*: 0.85; $R^2$: < 0.01). Furthermore, the
correlation observed in the north between $\Omega_A$ and $O_2$ is especially
strong when deep Channel\textcolor{red}{\sout{s}} \textcolor{blue}{waters} (*Deep* model: *p-value*: < 0.01; $O_2$: 0.20; $R^2$:
0.97) are separated from shallower regions (*Shallow* model: *p-value*: < 0.01;
$O_2$: 0.07; $R^2$: 0.96), with $\Omega_A$ increasing more steeply with
increasing $O_2$ in \textcolor{blue}{the} deep Channels. This stronger correlation is likely caused by the lack of gas exchange with the atmosphere or the surface mixed layer and the importance of metabolic $CO_2$ in deep waters, some of which may have mean ages of up to 20 years [@mucci2011]. In the Northern Gulf, the acidification \textcolor{red}{\sout{is following}} \textcolor{blue}{trends with} depth and distance from the Cabot Strait along the Laurentian, Anticosti and Esquiman Channels due to the progressive oxygen depletion and metabolic carbon dioxide accumulation [@mucci2011].

There are likely different dynamics at play in the Southern Gulf, such as waters with lowered alkalinity and salinity caused by the \textcolor{blue}{surface, seaward-flowing Gaspé Current} \textcolor{red}{\sout{St. Lawrence plume}}. \textcolor{red}{\sout{Such}} \textcolor{blue}{These} water\textcolor{blue}{s} \textcolor{red}{\sout{masses}} \textcolor{red}{\sout{are}} have lower buffering capacities to decreasing pH and $\Omega_{A}$. \textcolor{red}{\sout{Also}} \textcolor{blue}{Furthermore}, the accumulation of metabolic carbon dioxide produced by rapid oxic degradation of organic matter in the sandy and permeable sediments of the Magdalen Shelf or \textcolor{red}{\sout{through}} \textcolor{blue}{in} the \textcolor{red}{\sout{St. Lawrence plume}} \textcolor{blue}{Gaspé current} could decrease $\Omega_{A}$ in the Southern Gulf (Starr and Chassé, *personal communication*).

We thus divided the \textcolor{red}{\sout{analyses}} \textcolor{blue}{and their analysis} between the Southern and Northern St. Lawrence. The St. Lawrence was divided \textcolor{red}{\sout{using}} \textcolor{blue}{along} the 350m isobath on the southern slope of the Laurentian Channel, from the Cabot Strait to the tip of the Gaspé Peninsula. \textcolor{blue}{The} $\Omega_A$ distribution in the Southern St. Lawrence was modeled
using the exponential kriging model.

We predicted $\Omega_A$ using $O_2$ for the Northern St. Lawrence. Whereas
there are few published datasets on $\Omega_{A}$ in the St.
Lawrence, there is a robust time-series on oxygen concentration\textcolor{blue}{s} in the water column \textcolor{red}{\sout{from}} \textcolor{blue}{available through}
the Department of Fisheries and Oceans’ (DFO) Atlantic Zone
Monitoring Program [AZMP; @galbraith2018]. We used oxygen concentration data
collected between 2013 and 2017
[@blais2019] and interpolated oxygen % saturation using cokriging with depth as a covariable, as done in @dutil2011. Bathymetric data come from @dutil2011
and have a 100 $km^2$ resolution. For more details, refer to the hypoxia \textcolor{blue}{data layer}
description. The *Deep* model was used to transform $O_2$ values to $\Omega_A$
in grid cells with depth > 350m, while the *Shallow* model was
applied to cells with depth < 350m.

Based on the wide range of $\Omega_A$ levels at which organism\textcolor{blue}{s} are negatively
affected (see above), we built an index of acidification stress ($A_s$)
ranging between 0 and 1 that slowly \textcolor{red}{\sout{begins increasing}} \textcolor{blue}{increases} as $\Omega_A$ \textcolor{blue}{decreases to} $\approx$
$2$, increases more rapidly close to $\Omega_A$ $= 1$ and reaches its peak at
$\Omega_A$ $\approx$ $0.5$:

@@

The data used to characterize hypoxia
\textcolor{red}{\sout{uses oxygen saturation in \% and}}
come from
the Department of Fisheries and Oceans’ (DFO) Atlantic Zone Monitoring Program
[AZMP; @galbraith2018] survey\textcolor{red}{\sout{s}} in late spring and fall
\textcolor{blue}{of 2013 to 2017}, as well as from \textcolor{red}{\sout{annual}}
DFO’s \textcolor{blue}{annual} multispecies surveys for the northern Gulf in August and for the Magdalen
Shallows in September. We provide a brief summary of data and methods to
describe hypoxia in this document. For more details, refer to @blais2019.
